

(`r nrdrs %>% filter(tdomain == 'Bacteria') %>% distinct(genome_accno) %>% nrow()`) and only a few are archaeal
(`r nrdrs %>% filter(tdomain == 'Archaea') %>% distinct(genome_accno) %>% nrow()`).
`r  round(nrdrs %>% filter(tdomain == 'Bacteria') %>% distinct(genome_accno) %>% nrow() / taxa %>% filter(tdomain == 'Bacteria') %>% distinct(genome_accno) %>% nrow() * 100)`% 
of bacterial species, but only
`r  round(nrdrs %>% filter(tdomain == 'Archaea') %>% distinct(genome_accno) %>% nrow() / taxa %>% filter(tdomain == 'Archaea') %>% distinct(genome_accno) %>% nrow() * 100)`% 
of archaeal species code for NrdR.

WRITE SOMETHING ABOUT WHAT YOU FIND IN THE PLOTS. REFER TO PLOTS AND TABLES WITH (check source code) `r figr('genomes-with-nrdr', T, type = 'Figure')`.

```{r nrdrs-per-taxon-level-func}
# Ghada simplifies so function only takes domain. Remember to change name of function.
nrdrs_per_taxon_level <- function(domain, level) {
  nrdrs %>%
    filter(tdomain == domain) %>%
    group_by_at(level) %>%
    distinct(genome_accno) %>% 
    summarise(nrdr_pres_count = n(), .groups = 'drop') %>%
    full_join(
      taxa %>%
        filter(tdomain == domain) %>%
        group_by_at(level) %>%
        summarise(genome_count = n(), .groups = 'drop'),
      by = level
    ) %>%
    replace_na(list(nrdr_pres_count = 0)) %>%
    mutate(proportion = nrdr_pres_count / genome_count) 
  
  # Ghada finishes the plot...
}
```

```{r bacterial-nrdrs-per-phylum}
nrdrs_per_taxon_level('Bacteria', 'tphylum')
```
```{r Bacteria-Phymum-function}
domain <- 'Bacteria'
level <- 'tphylum'
```

```{r Archaea-Phymum-function}
domain <- 'Artchaea'
level <- 'tphylum'
```

```{r genomes-with-nrdr, fig.cap = "Genomes with NrdR in main taxonomic groups."}
# Call function according to the desired domain/taxon combination

nrdr_by_level <- nrdrs %>%
  filter(tdomain == domain) %>%
  group_by_at(level) %>%
  distinct(genome_accno) %>% 
  summarise(nrdr_pres_count = n(), .groups = 'drop') %>%
  full_join(
    taxa %>%
      filter(tdomain == domain) %>%
      group_by_at(level) %>%
      summarise(genome_count = n(), .groups = 'drop'),
    by = level
  ) %>%
  replace_na(list(nrdr_pres_count = 0)) %>%
  mutate(proportion = nrdr_pres_count / genome_count) 
```

```{r NrdR-by-taxon-plot}
# ###Plotting 
nrdr_by_level %>% 
  filter(genome_count > 1) %>%
  ggplot(aes_string(x=level, y="genome_count", fill="proportion")) +
  geom_bar(stat="identity") +
  scale_fill_gradient(low = "pink", high = "purple", na.value = NA) +
  ylab('Number of genomes') +
  xlab(sprintf('%s %s', domain, level)) +
  scale_y_log10() +
  guides(fill=guide_legend(title="Proportion of genomes with NrdR")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

# Results
Bacterial Phyla with no nrdr or nrdr_genomes proportion < 0.01, exclusing single genome phyla, are:
`r genomes-with-nrdr %>% filter(proportion < 0.01 & genome_count > 1)` 

Archaeal Phyla with nrdr arre:
`r genomes-with-nrdr %>% filter(proportion > 0.7 & genome_count > 1)`

```{r nrdr-with-single-rnr, fig.cap = "NrdR presence with single RNR"}
nrdr_srnr <- rnrs %>% 
  group_by(genome_accno) %>% 
  mutate (rnrs_by_genome = n()) %>%
  filter(rnrs_by_genome == 1) %>%
  inner_join(nrdrs %>% transmute(genome_accno, NrdR_accno = accno), by = 'genome_accno') %>%
  distinct(genome_accno, profile, accno, tdomain, tphylum, tclass, torder, tfamily, tgenus, tspecies) 
    
# ##Proportion of nrdr with single RNR in the genome (tibble output)
# ##CALL domain/Taxon chunk

nrdr_srnr_prop <- nrdr_srnr %>%
  filter(tdomain == domain) %>%
  group_by_at(level) %>%
  mutate(count = n()) %>%
  inner_join(
    taxa %>%
      group_by_at(level)%>%
      summarize(genome_by_level = n()),
    by = level
  ) %>%
  mutate(proportion = count / genome_by_level) 

nrdr_srnr_prop %>% 
  distinct_at(level) %>%
  inner_join(nrdr_srnr_prop %>% select(proportion) %>% distinct(proportion), by = level)
```

```{r singleRNR-NrdR-genomes-plot}
 ### Plot proportion of NrdR with single RNR on genome by taxonomic level
sublevel <- '' # DL: What should this be set to?
nrdr_srnr_prop %>% 
  filter(tdomain == domain) %>%
  filter(genome_by_level > 1) %>%
  distinct(tphylum, count, tdomain) %>%
  ggplot(aes_string(x="tdomain", y="count", fill=level)) +
  geom_bar(stat="identity") +
  ylab('Number of genomes with NrdR and a single RNR') +
  xlab(sprintf('NrdR with single RNR in %s', domain)) +
  guides(fill=guide_legend(title=sprintf("Distribution by %s", sublevel))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

```{r nrdr-with-multiple-rnrs, fig.cap = "NrdR presence with multiple RNR"}
nrdr_mrnr <- rnrs %>% 
  group_by(genome_accno) %>% 
  mutate (rnrs_by_genome = n()) %>%
  filter(rnrs_by_genome > 1) %>%
  inner_join(nrdrs %>%  transmute(genome_accno, NrdR_accno = accno),  by = 'genome_accno') %>%
  distinct(genome_accno, profile, accno, tdomain, tphylum, tclass, torder, tfamily, tgenus, tspecies) 

# Proportion of nrdr with multiple RNR in the genome (tibble output)
####Call domain/taxon function

nrdr_mrnr_prop <- nrdr_mrnr %>%
  filter(tdomain == domain) %>%
  group_by_at(level) %>%
  mutate(count = n()) %>%
  inner_join(taxa %>% group_by_at(level)%>% summarize(genome_by_level = n()), by = level) %>%
  mutate(proportion = count / genome_by_level) 

nrdr_mrnr_prop %>% 
  distinct_at(level) %>%
  inner_join(nrdr_mrnr_prop %>% select(proportion) %>% distinct(proportion), by = level)
```  

```{r multipleRNR-NrdR-genomes-plot}
# ##Plot proportion of NrdR with multiple RNR on genome by taxonomic level
nrdr_mrnr_prop %>% 
  filter(tdomain == domain) %>%
 filter(genome_by_level > 1) %>%
  distinct(tphylum, proportion) %>%
  ggplot(aes_string(x="tphylum", y="proportion", fill=level)) +
  geom_bar(stat="identity") +
  ylab('Number of genomes with NrdR and multiple RNR') +
  xlab(sprintf('NrdR with multiple RNR in %s', domain)) +
  guides(fill=guide_legend(title=sprintf("%s", sublevel))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```
  
```{r nrdr-with-different-classes, fig.cap = "NrdR presence with multiple RNR/different classes"}
# Proportion of nrdr with multiple RNR in the genome (tibble output) but from different RNR classes 
# ###Call domain/taxon function

nrdr_drnr <- nrdr_mrnr %>%
  inner_join(hmm_profiles %>% select(profile, pclass), by = 'profile') %>%
  filter(tdomain == domain) %>%
  group_by(genome_accno) %>%
  mutate(pcount= n()) %>%
  ungroup %>%
  distinct(genome_accno, tphylum, pclass, pcount) %>%
  pivot_wider(names_from = pclass, values_from = pcount)
  nrdr_drnr = nrdr_drnr %>% mutate(na_count = rowSums(is.na(nrdr_drnr))) %>%
  filter(na_count < 2)

# Occurence of Nrdr with each rnr class combination
# DL: This can be simplified!
nrdr_drnr %>%
  filter(is.na(NrdD)) %>%
  group_by_at(level) %>%
  mutate(NrdAJ = n()) %>%
  distinct(tphylum, NrdAJ) %>%
  full_join(
    nrdr_drnr %>%
      filter(is.na(NrdJ)) %>%
      group_by_at(level) %>%
      mutate(NrdAD = n()) %>%
      distinct(tphylum, NrdAD),
    by = level
  ) %>%
  full_join(nrdr_drnr %>%
               filter(is.na(NrdA)) %>%
               group_by_at(level) %>%
               mutate(NrdJD = n()) %>%
              distinct(tphylum, NrdJD),
             by = level) %>%
  inner_join(taxa %>%
              group_by_at(level)%>%
              summarize(genome_by_level = n()),
            by = level) %>%
  mutate(NrdAJ_prop = NrdAJ / genome_by_level, NrdAD_prop = NrdAD / genome_by_level, NrdDJ_prop = NrdJD / genome_by_level) 
```
  
```{r nrdr-in-archaea-w-nrda, fig.cap = "NrdR in Archaea with class I RNR"}

# NrdR+NrdA combination existed in one Archaeal class in each of the three phyla that had this combination (i.e. looking at phylum level, no variability by tclass was found, so we checked by torder)
sublevel <- 'torder'

nrdr_nrda <- nrdrs %>%
  filter(tdomain == 'Archaea') %>%
  select(genome_accno, level, sublevel, tspecies) %>%
  semi_join(rnrs %>% filter(pclass == 'NrdA'), by = 'genome_accno') %>%
  group_by_at(level) %>%
  mutate(count= n()) %>%
  inner_join(taxa %>% group_by_at(level)%>% summarize(genome_by_level = n()), by = level) %>%
  mutate(proportion = count/genome_by_level) 

# ##Proportion of genomes with NrdR and NrdA by phylum
nrdr_nrda %>% distinct(tphylum, count, proportion)
```

```{r NrdR-classI-plot}
# ##Plotting NrdR + class I RNR combination by taxonomic level
nrdr_nrda %>% 
  distinct(tphylum, torder, count, genome_by_level, proportion) %>%
  # filter(genome_by_level > 1) %>%
  ggplot(aes_string(x=level, y="proportion", fill=sublevel)) +
  geom_bar(stat="identity") +
  ylab(sprintf('Proportion of %s genomes with NrdR and class I RNR', domain)) +
  xlab(sprintf('%s with NrdR and class I RNR in %s', level, domain)) +
  guides(fill=guide_legend(title=sprintf("%s", sublevel))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```
  
```{r nrdr-w-rnr-on-same-strand, fig.cap = "Presence of NrdR with RNR on the same strand/operon"}

```
  
```{r nrdr-paralogs, fig.cap = "NrdR paralogs"}

```

