---
title: "Overview of RNR proteins in GTDB *species*"
author: "daniel.lundin@dbb.su.se & ghada.nouaria@dbb.su.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: true
    code_folding: hide
  pdf_document:
    fig_caption: yes
    fig_height: 9
    fig_width: 8
    number_sections: yes
    toc: yes
---

```{r setup, echo=F, cache = FALSE}
knitr::opts_chunk$set(echo=F, fig.path='figures/', cache = TRUE)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries, message=F, cache = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(feather))
#suppressPackageStartupMessages(library(data.table))
#suppressPackageStartupMessages(library(dtplyr))
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(kfigr))
suppressPackageStartupMessages(library(knitr))
```

```{r constants}
```

```{r read-data}
dbsources    <- read_feather("data/pfitmap-gtdb.dbsources.feather")
hmm_profiles <- read_feather("data/pfitmap-gtdb.hmm_profiles.feather")
taxa         <- read_feather("data/pfitmap-gtdb-rep.taxa.feather")
accessions   <- read_feather("data/pfitmap-gtdb-rep-RNRs.accessions.feather")
### domains      <- read_feather("data/pfitmap-gtdb-rep-RNRs.domains.feather")
proteins     <- read_feather("data/pfitmap-gtdb-rep-RNRs.proteins.feather")
### sequences    <- read_feather("data/pfitmap-gtdb-rep-RNRs.sequences.feather")

# Convenience tables and numbers
rnr_alphas <- hmm_profiles %>% filter(pclass %in% c('NrdA', 'NrdD', 'NrdJ'))

major_phyla <- taxa %>% 
  count(tdomain, tphylum, name = "n_genomes") %>% 
  group_by(tdomain) %>% top_n(11, n_genomes) %>% ungroup() %>%
  mutate(major_phylum = tphylum)
major_phyla <- major_phyla %>% 
  union(
    taxa %>% anti_join(major_phyla, by = c('tdomain', 'tphylum')) %>%
      count(tdomain, tphylum, major_phylum = NA, name = "n_genomes")
  ) %>%
  mutate(major_phylum = forcats::fct_explicit_na(major_phylum, na_level = 'Other phyla'))

genomes_wo_rnr_alpha <- taxa %>% 
  select(genome_accno, tdomain:tspecies) %>%
  anti_join(
    rnr_alphas %>% inner_join(proteins, by = 'profile') %>%
      inner_join(accessions, by = 'accno'),
    by = "genome_accno"
  ) 

n_rnr_species <- rnr_alphas %>%
  inner_join(proteins %>% select(profile, accno), by = 'profile') %>% 
  inner_join(accessions %>% distinct(accno, genome_accno), by = 'accno') %>% 
  distinct(genome_accno) %>% count() %>% pull(n)
```

# Data

The data in this document was based on a search for RNR proteins in  `r nrow(taxa)` 
species in the  `r dbsources %>% pull(version)`  release of the GTDB database. The
GTDB database only contains Archaea and Bacteria.

# RNR presence in genomes

Of a total of `r nrow(taxa)` species,
`r n_rnr_species` contained at least one catalytic subunit RNR protein, i.e.
`r round(n_rnr_species/nrow(taxa)*100)`%. 
Genome completeness does not seem to be a major factor in explaining the absence of RNR alpha subunits
(`r figr('missing-rnr-completeness', T, type = 'Figure')`).
Nor does different requirements for how much of the HMM model is covered by alignments appear to influence
the number of genomes with at least one RNR alpha subunit
(`r figr("num-genomes-hmm-coverage", T, type = "Figure")`).

In many phyla around 10-30% of species lack identified RNR alpha subunits
(`r figr("no-rnr-alpha-phylum-dist", T, type = "Figure")`).

```{r missing-rnr-completeness, fig.height = 4, fig.cap = '**Number of species representative GTDB genomes missing any RNR alpha subunit as a function of genome completeness.**'}
taxa %>% transmute(
  genome_accno, 
  tdomain, 
  completeness = as.numeric(checkm_completeness),
  compl_interval = ceiling(completeness / 5) * 5
) %>%
  anti_join(
    rnr_alphas %>% inner_join(proteins, by = 'profile') %>% inner_join(accessions, by = 'accno'),
    by = 'genome_accno'
  ) %>%
  count(tdomain, compl_interval) %>%
  ggplot(aes(x = compl_interval, y = n, fill = tdomain)) +
  geom_col() +
  scale_fill_discrete('Domain') +
  ylab('N. genomes without RNR') + xlab('Completeness percent intervall')
```

```{r num-genomes-hmm-coverage, fig.height = 4, fig.cap = "**Number of genomes with at least one RNR catalytic subunit as a function of different degrees of HMM model coverage.** Although the number of genomes appear to decrease radically between 0.85 and 0.9, note that the y-axis scale is very short and that it's actually only about 250 species that lose their RNRs between these levels."}
tibble(min_hmm_coverage = c(0.25, 0.5, 0.6, 0.7, 0.75, 0.8, 0.85, 0.9, 0.95, 1.00)) %>%
  mutate(
    n_genomes = purrr::map(
      min_hmm_coverage,
      function(cov) {
        rnr_alphas %>% 
          inner_join(proteins, by = 'profile') %>%
          filter(plen / hmmlen >= cov) %>%
          inner_join(accessions, by = 'accno') %>%
          count()
      }
    )
  ) %>%
  unnest(n_genomes) %>%
  ggplot(aes(x = min_hmm_coverage, y = n)) +
  geom_line() +
  xlab('Minimum HMM model coverage') +
  ylab('Number of genomes with at least on RNR alpha subunit')
```

```{r no-rnr-alpha-phylum-dist, fig.height = 5, fig.cap = "**Species missing RNR alpha subunits per phylum.** The top two panels show the number of species that do not encode any identified RNR alpha subunit, while the bottom two show the proportion of species per phylum that do not encode RNR alphas."}
genomes_wo_rnr_alpha %>%
  count(tdomain, tphylum) %>%
  inner_join(major_phyla, by = c('tdomain', 'tphylum')) %>%
  group_by(tdomain, major_phylum) %>% summarise(n = sum(n), n_genomes = sum(n_genomes), .groups = 'drop') %>%
  transmute(tdomain, major_phylum, `Proportion of genomes` = n/n_genomes, `Number of genomes` = n) %>%
  pivot_longer(c(`Proportion of genomes`, `Number of genomes`), names_to = 'type', values_to = 'meas') %>%
  ggplot(aes(x = major_phylum, y = meas)) +
  geom_col() +
  facet_wrap(type~tdomain, scales = 'free') +
  coord_flip() +
  xlab("Phylum") +
  ylab("Number or proportion of species missing RNR alpha subunits")
```

